---
title: "Day3: R for data visualization"
author: "Wenbin Guo"
date: "2024-04-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Welcome to the QCBio workshop W3: `Introduction to R and data visualization`

# Day3 outline
For today's workshop practice, we will cover the following topics
- getting started with ggplot2 
- basic plots: scatter plot, line plot, bar plot, histogram, boxplot
- customization a plot
- statistical testing/modeling


# load packages
```{r}
# install/load package
if(!require(tidyverse)){
  install.packages("tidyverse")
}

library(tidyverse)
```


# Recap
## variables
### variable types
```{r}
x = c(0, 1.5, -2)
y = c("low", "medium", "high")
```

### factors
check the results of following 2 lines about factors, what do you find?
```{r}
y1 = factor(y)
y2 = factor(y, levels = c("low", "medium", "high"))
```


# ggplot2
## build plot layer by layer
```{r}
base = ggplot(data=mpg, aes(x=hwy, y=displ))

base

base + 
  geom_point(aes(colour = drv))

base + 
  geom_point(aes(colour = drv)) +   
  geom_smooth(method = "lm") 


base + 
  geom_point(aes(colour = drv)) +   
  geom_smooth(method = "lm") +   
  theme_minimal() +   
  ggtitle("Relationship between variables in mpg dataset", 
          subtitle = "hwy vs displ")
```


## Aesthetic mapping
```{r}
base = ggplot(data = mpg)

base + geom_point(aes(x = displ, y = hwy, color = drv))
base + geom_point(aes(x = displ, y = hwy, size  = drv))
base + geom_point(aes(x = displ, y = hwy, alpha = drv))
base + geom_point(aes(x = displ, y = hwy, shape = drv))
```


### mapping vs setting
```{r}
ggplot(mpg) + geom_point(aes(displ, hwy, colour = "blue")) # mapping, will treat as a new variable, assign the default color
ggplot(mpg) + geom_point(aes(displ, hwy), colour = "blue") # setting, will use the designed color
```

**Exercise**: build a plot that contains the following elements
- set `hwy` as x axis, and `cty` as `y` axis
- set point shapes and color according to `cyl`
- set theme as `theme_bw()`
- add title `scatter plot` and subtitle `cty.vs.hwy`

> Hint: the variable types matter!


## Geometry
### basic types
```{r}
base = ggplot(data = mpg)

# blank
base + geom_blank() 

# lines
base + aes(displ, hwy) + 
  geom_vline(xintercept = 3) + geom_hline(yintercept = 25) 

base + aes(displ, hwy) + 
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "blue") 


# line vs path 
base + aes(displ, hwy) + 
  geom_point() + geom_line()

base + aes(displ, hwy) + 
  geom_point() + geom_path()


# region
base + aes(displ, hwy) + 
  geom_rect(xmin = 2.5, xmax=3.5, ymin = 25, ymax = 30, fill = "grey", color = "red")
```


### single variable
```{r}
base + geom_bar(aes(x = class))     # bar plot
base + geom_histogram(aes(x = hwy)) # histogram
base + geom_density(aes(x = hwy))   # density
```

**Exercise**: plot the barplot in ascending order (bar height increases from left to right)


### two variables
#### continous
```{r}
base + geom_point(aes(x = displ, y =hwy)) # scatter plot

base + 
  geom_point(aes(x = displ, y =hwy)) + 
  geom_smooth(aes(x = displ, y =hwy), method = "lm") # scatter plot with regression line
```

#### dicrete 
```{r}
base + geom_boxplot(aes(x = drv, y = hwy))  # boxplot

base + geom_violin(aes(x = drv, y = hwy))   # violin

base + geom_point(aes(x = drv, y = hwy))    # treat it as scatter plot
base + geom_jitter(aes(x = drv, y = hwy))   # jitter
```


### heatmap (using geom_tile)
```{r}
ht_data  = expand.grid(X = 1:10, Y = 1:10)
ht_data$Value = runif(100, min = 1, max = 100)

ggplot(ht_data) + 
  geom_tile(aes(x = X, y = Y, fill = Value))
```

Note: There are more specialized packages such as `pheatmap`, `complexHeatmap`, `ggheatmap` that can be used to make fancy heatmaps


## Scale
**Exercise**: make the following boxplot to have color "red", "blue", "green"
```{r}
ggplot(mpg) +
  geom_point(aes(x = drv, y = hwy), position = "jitter") +
  geom_boxplot(aes(x = drv, y = hwy))
```


## Facet
```{r}
base = ggplot(data = mpg)

base + geom_point(aes(displ, hwy, colour = drv)) + facet_null()

base + geom_point(aes(displ, hwy, colour = drv)) + facet_wrap(~class)

base + geom_point(aes(displ, hwy, colour = drv)) + facet_wrap(~class, scales = "free_x")

base + geom_point(aes(displ, hwy, colour = drv)) + facet_grid(year~class, scales = "free")
```


## plot title, subtitle, caption and tag
```{r}
plot1 = base + 
  geom_point(aes(x = displ, y = hwy, color = drv))


plot1 + labs(title = "mytitle", 
            subtitle = "my subtitle", 
            caption = "this is a caption",
            tag = "a)")
```


## theme
```{r}
# check the built in themes

plot1
plot1 + theme_bw()
plot1 + theme_classic()
plot1 + theme_dark()
plot1 + theme_grey()
plot1 + theme_light()
plot1 + theme_linedraw()
plot1 + theme_minimal()
plot1 + theme_test()
plot1 + theme_void()
```


## save plot
```{r}
ggsave("plot1.png", plot1, width = 5, height = 5, dpi = 300)

pdf("plot1.pdf", width = 5, height = 5)
print(plot1)
dev.off()
```


# Statistical testing and modeling
## t-test
```{r}
t.test(mpg %>% dplyr::filter(drv == "f") %>% dplyr::select(hwy), 
       mpg %>% dplyr::filter(drv == "4") %>% dplyr::select(hwy))
```


## correlation
```{r}
cor(mpg$hwy, mpg$cty)
cor.test(mpg$hwy, mpg$cty)
```


## linear model
```{r}
lm1 = lm(hwy ~ cty, data = mpg)
summary(lm1)
```


# 2048 in R
check the `2048.R` code and see if you can understand its logic

