---
title: "Day2: R advanced topics"
author: "Wenbin Guo"
date: "2024-04-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Welcome to the QCBio workshop W3: `Introduction to R and data visualization`

# Outline
## Day2 outline
For today's workshop practice, we will cover the following topics
- Flow control, loop, function, packages
- Data wrangling skills


## Tips on Rmarkdown usage in Rstudio:
- Ctrl/Cmd + Alt + i: insert a code chunk
- Ctrl/Cmd + Shift + c: comment/uncomment the code

For more information about the Rmarkdown usage, please refer to the following reference:
- [cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
- [R markdown cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/)
- [keyboard shortcuts](https://bookdown.org/yihui/rmarkdown-cookbook/rstudio-shortcuts.html)


# Flow control
## if-else
```{r}
x = 10

if(is.nan(x))
{
message("x is missing")
} else if(is.infinite(x))
{
message("x is infinite")
} else if(x > 0)
{
message("x is positive")
} else if(x < 0)
{
message("x is negative")
} else
{
message("x is zero")
}
```

## switch
```{r}
day_of_week_if_else <- function(day) {
  if (day == 1) {
    result <- "Monday"
  } else if (day == 2) {
    result <- "Tuesday"
  } else if (day == 3) {
    result <- "Wednesday"
  } else if (day == 4) {
    result <- "Thursday"
  } else if (day == 5) {
    result <- "Friday"
  } else if (day == 6) {
    result <- "Saturday"
  } else if (day == 7) {
    result <- "Sunday"
  } else {
    result <- "Invalid day"
  }
  return(result)
}
```

```{r}
day_of_week_switch <- function(day) {
  result <- switch(as.character(day),
                   "1" = "Monday",
                   "2" = "Tuesday",
                   "3" = "Wednesday",
                   "4" = "Thursday",
                   "5" = "Friday",
                   "6" = "Saturday",
                   "7" = "Sunday",
                   "Invalid day")
  return(result)
}
```


```{r}
day_of_week_if_else(3)
day_of_week_switch(3)
```

## one line if-else
```{r}
a = 3
b = if(a > 0) "positive" else "non-positive"
b
```



## vectorized if-else
```{r}
x = c(10, -5)

## The following code is expected to have error because x is a vector
if(x>0){
  print("positive")
} else {
  print("non-positive")
}
```


```{r}
ifelse(x>0, "positive", "non-positive")
```

# loop
## for loop
```{r}
for(i in 1:5){
  j = i^2
  print(paste0("j=", j))
}
```

## while loop
```{r}
i = 1
while(i <= 5){
  j = i^2
  print(paste0("j=", j))
  i = i + 1
}
```


## repeat loop
```{r}
i = 1
repeat{
  j = i^2
  print(paste0("j=", j))
  i = i + 1
  if(i > 5){
    break
  }
}

```

also, check break/next if you want to break/continue the loop


## apply family
```{r}
my_list = c(list(a = 1, b = 2), list(3, 4))
lapply(my_list, sqrt)
```

```{r}
sapply(my_list, sqrt)
```


```{r}
group = rep(1:2,each =3)
value = c(1,2,3,4,5,6)
tapply(value, group, sum)
```


```{r}
mat = matrix(1:9, nrow = 3)
apply(mat, 1, sum)
```



# function
## create
```{r}
hypotenuse <- function(x, y)
{
sqrt(x ^ 2 + y ^ 2)
}


hypotenuse2 <- function(x = 5, y = 12)
{
sqrt(x ^ 2 + y ^ 2)
}

h = function(x){
  x * y
}
```


## call
```{r}
# the following code is expected to have error because arguments are not provided
hypotenuse()
```


```{r}
hypotenuse2()
```

```{r}
# the following code is expected to have error because y is not provided
h(2)
```


## function body
funtion body can be retrieved by `body()` function
```{r}
body(h)
```

# packages
```{r}
# install/load package
if(!require(tidyverse)){
  install.packages("tidyverse")
}

library(tidyverse)
```


# File I/O
## read in data
```{r}
# read in data, might need to adjust the path
tcga_sample = read.table("day2/tcga_sample.csv", sep = ",", header = TRUE)
```


## write data
```{r}
saveRDS(tcga_sample, "day2/tcga_sample.rds")
tcga_sample2= readRDS("day2/tcga_sample.rds")
```



# Data wrangling basics
## load data mpg
Fuel economy data from 1999 to 2008 for 38 popular models of cars. 
```{r}
# load mpg
data(mpg)

# check the first several lines
head(mpg)
```

meaning of variables:
- `cty` and `hwy` record miles per gallon (mpg) for city and highway driving.
- `displ` is the engine displacement in liters.
- `drv` is the drivetrain: front wheel (f), rear wheel (r) or four wheel (a).
- `model` is the model of car. 
- `class` is a categorical variable describing the type of car: two seater, SUV, compact, etc.



## variables
### variable types
```{r}
x = c(0, 1.5, -2)
y = c("low", "medium", "high")
```

### factors
check the results of following 2 lines about factors, what do you find?
```{r}
y1 = factor(y)
y2 = factor(y, levels = c("low", "medium", "high"))
```


## data frame
the `mpg` dataset is a typical example of data frame

### create a new data frame
```{r}
df = data.frame(x = c(1,2,3), y = c("low", "medium", "high"))
df2= data.frame(x = c(1,2,3), y0 = c("apple", "banana", "organge"))

head(df)
```

## Manipulate observations
### filter
Filtering is useful for extracting outliers or remove abnormal data points (such as missing values). 

Filtering is also a useful trick when we dealing with big datasets. By filtering, we can explore a small part of the data before we scale up to the whole dataset. This could save us time when building the prototype of processing and visualization pipeline 

```{r}
# singlecomparison operators
filter(mpg, model == "a4")

# with logical operators (compound condition)
filter(mpg, model == "a4" & cyl==4)
```


### arrange
```{r}
# default ascending
arrange(mpg, hwy)

# descending
arrange(mpg, desc(hwy))
```


### merge by rows
use `bind_rows()` to merge 2 dataframe by rows
```{r}
mpg_df1 = filter(mpg, manufacturer == "audi")
mpg_df2 = filter(mpg, manufacturer == "ford")

bind_rows(mpg_df1, mpg_df2)
```


**Exercise**: extract observations whose `cty` value is among the top 10

**Exercise**: extract observations whose manufacturer is `audi` or `ford`, filter by `cty > 18` , and then arrange the resulting data frame by `cty` in descending order

**Exercise**: run `bind_rows(df,df2)`, what would you find?




## Manipulate variables
### select
```{r}
mpg_df3 = dplyr::select(mpg, manufacturer, model, year, displ)
mpg_df4 = dplyr::select(mpg, manufacturer, model, year, hwy, cty)

mpg_df3
mpg_df4
```


### mutate
```{r}
# get a ratio of hwy, cty
mutate(mpg_df4, ratio = hwy/cty)
```


### merge by columns
**Exercise** Try to use `inner_join()`, `full_join()`, `left_join()`, `right_join()` to merge `mpg_df3` and `mpg_df4` by columns




## Reshape data
convert the data to long/wide format using function `pivot_longer()` and `pivot_wider()`in tidyr
```{r}
mpg_df4_long = pivot_longer(mpg_df4, cols = c("hwy", "cty"), names_to = "type", values_to = "mile_per_gallon")
mpg_df4_long
```

Exercise, try to convert the `mpg_df4_long` to the wide format




## Group & Summarize 
group by manufacturer and year, and compute the mean of `hwy` and `cty`
```{r}
mpg_group = group_by(mpg, year, manufacturer)
mpg_summary = summarise(mpg_group, hwy_mean = mean(hwy), cty_mean = mean(cty))
mpg_summary
```




## Pipe
use pipe to chain the functions together
```{r}
mpg %>% group_by(year, manufacturer) %>% 
  summarise(hwy_mean = mean(hwy), cty_mean = mean(cty))
```

**Exercise** select the observation that has the smallest `cty` for each year.
